---
title: "Penalized Estimation of Cross-Loadings and Unique Covariances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Penalized Estimation of Cross-Loadings and Unique Covariances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r}
library(plavaan)
library(lavaan)
data(PoliticalDemocracy)
```

## Penalize cross-loadings

### Two-factor CFA model

```{r}
mod0 <- "
  ind60 =~ x1 + x2 + x3
  dem60 =~ y1 + y2 + y3 + y4
  ind60 ~~ dem60
"
fit0 <- cfa(mod0, data = PoliticalDemocracy, std.lv = TRUE)
```

### Two-factor EFA model (unidentified)

```{r}
mod <- "
  ind60 =~ x1 + x2 + x3 + y1 + y2 + y3 + y4
  dem60 =~ x1 + x2 + x3 + y1 + y2 + y3 + y4
  ind60 ~~ ind60
"
fit <- cfa(mod, data = PoliticalDemocracy, std.lv = TRUE, do.fit = FALSE)
```

### Two-factor EFA model with penalized cross-loadings

The cross-loadings are the parameters 4 to 10 in the parameter table (see the `free` column).

```{r}
parTable(fit)
```

```{r}
pefa_fit <- penalized_est(
    fit,
    w = .03,
    pen_par_id = 4:10,
    test = "Chisq"
)
summary(pefa_fit)
```

Fit indices can be obtained directly from the penalized fit. Fit evaluation is experimental and disabled by default, so the fit above was created with `test = "Chisq"`; with the default `test = "none"`, `fitmeasures()` is unavailable and `summary()` shows no chi-square test. When enabled, `fitmeasures()` refits the model with all parameters frozen at the penalized estimates and reports the indices at the effective degrees of freedom (11.97 here, versus the nominal 6), and an experimental notice is shown. For the under-identified models later in this vignette (negative nominal df), the effective df is the meaningful value.

```{r}
fitmeasures(pefa_fit, c("chisq", "df", "cfi", "rmsea"))
```

## Penalize Cross-loadings and Unique Covariances

### Two-factor EFA model with unique covariances

```{r}
mod2 <- "
  ind60 =~ x1 + x2 + x3 + y1 + y2 + y3 + y4
  dem60 =~ x1 + x2 + x3 + y1 + y2 + y3 + y4
  ind60 ~~ ind60
  x1 ~~ x2 + x3 + y1 + y2 + y3 + y4
  x2 ~~ x3 + y1 + y2 + y3 + y4
  x3 ~~ y1 + y2 + y3 + y4
  y1 ~~ y2 + y3 + y4
  y2 ~~ y3 + y4
  y3 ~~ y4
"
fit2 <- cfa(mod2, data = PoliticalDemocracy, std.lv = TRUE, do.fit = FALSE)
```

### Two-factor EFA model with penalized cross-loadings and unique covariances

The unique covariances are the parameters 15 to 35 in the parameter table (see the `free` column).

```{r}
parTable(fit2)
```

```{r}
pefa_fit2 <- penalized_est(
    fit2,
    w = .03,
    pen_par_id = c(4:10, 15:35)
)
summary(pefa_fit2)
```

The unique covariances were all estimated close to zero. The effective number of cross-loadings and unique covariances can be reported with `effective_df()`:

```{r}
effective_df(pefa_fit2)
```

So out of 28 penalized parameters, only about 1.6 (or close to 2) are effectively non-zero. The table also shows the effective model degrees of freedom (11.44), where the nominal df (-15) is negative because the model is under-identified.

## Penalize Cross-Loadings, Unique Covariances, and Difference in Loadings and Intercepts Across Time

First, the model without cross-loadings and concurrent unique covariances

```{r}
mod3 <- "
    ind60 =~ NA * x1 + x2 + x3
    dem60 =~ NA * l1 * y1 + l2 * y2 + l3 * y3 + l4 * y4
    dem65 =~ NA * l1 * y5 + l2 * y6 + l3 * y7 + l4 * y8
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
    ind60 ~~ 1 * ind60
    dem60 ~~ 1 * dem60
    dem65 ~~ NA * dem65
    ind60 ~ 0 * 1
    dem60 ~ 0 * 1
    dem65 ~ NA * 1
    x1 + x2 + x3 ~ NA * 1
    y1 ~ i1 * 1
    y2 ~ i2 * 1
    y3 ~ i3 * 1
    y4 ~ i4 * 1
    y5 ~ i1 * 1
    y6 ~ i2 * 1
    y7 ~ i3 * 1
    y8 ~ i4 * 1
    y1 ~~ y5
    y2 ~~ y6
    y3 ~~ y7
    y4 ~~ y8
"
fit3_base <- cfa(mod3, data = PoliticalDemocracy)
```

```{r}
# Lavaan example of Political Democracy
mod3_un <- "
    ind60 =~ NA * x1 + x2 + x3 + y1 + y2 + y3 + y4
    dem60 =~ NA * x1 + x2 + x3 + y1 + y2 + y3 + y4
    dem65 =~ NA * y5 + y6 + y7 + y8
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
    ind60 ~~ 1 * ind60
    dem60 ~~ 1 * dem60
    dem65 ~~ NA * dem65
    ind60 ~ 0 * 1
    dem60 ~ 0 * 1
    dem65 ~ NA * 1
    x1 + x2 + x3 + y1 + y2 + y3 + y4 ~ NA * 1
    y5 + y6 + y7 + y8 ~ NA * 1
    x1 ~~ x2 + x3 + y1 + y2 + y3 + y4
    x2 ~~ x3 + y1 + y2 + y3 + y4
    x3 ~~ y1 + y2 + y3 + y4
    y1 ~~ y2 + y3 + y4
    y2 ~~ y3 + y4
    y3 ~~ y4
    y1 ~~ y5
    y2 ~~ y6
    y3 ~~ y7
    y4 ~~ y8
"
fit3 <- cfa(
    mod3_un,
    data = PoliticalDemocracy,
    do.fit = FALSE,
    start = fit3_base
)
```

```{r}
pt3 <- parTable(fit3)
# Provide better starting values
pt3$start[c(4:10, 35:55)] <- 0
fit3_2 <- lavaan::cfa(
    pt3,
    data = PoliticalDemocracy,
    do.fit = FALSE
)
```

Parameter IDs:

- (Concurrent) Cross-loadings: 4 to 10
- (Concurrent) Unique covariances: 35 to 55
- Loadings across time: 11 to 18
- Intercepts across time: 27 to 34


```{r}
pefa_fit3 <- penalized_est(
    fit3_2,
    w = .03,
    pen_par_id = c(4:10, 35:55),
    pen_diff_id = list(
        loadings = rbind(11:14, 15:18),
        intercepts = rbind(27:30, 31:34)
    )
)
summary(pefa_fit3, standardized = TRUE)
```

`effective_df()` reports the effective number of cross-loadings and unique covariances that are non-zero, and the effective number of loadings and intercepts that differ across time:

```{r}
effective_df(pefa_fit3)
```

For a difference-penalty block, `npar_effective` is the number of columns (one shared invariance baseline per parameter) plus the effective number of non-invariant values. Here the `loadings` and `intercepts` rows (4.02 and 4.03, versus 4 baseline values each) indicate that the loadings and intercepts are effectively invariant across time, while only about 1.3 of the 28 cross-loadings and unique covariances are effectively non-zero.